---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

In this codebook, we will complete the instrusticons step by step in the given
order. According to the research pipepline, we will have some processing and
analytic code in this codebook.

## Phase 1: Parsing Data

As you can see from the directory we are working on, we have our dataset as a
ZIP file we will first unzip the file and gather the dataset from the new 
directory

```{r}
#First, let's learn our directory and set it to a variable
directory <- getwd()

#Let's check for the filename and append it to a variable and unzip that file.
datazip <-list.files(directory, pattern = ".zip")
unzip(datazip)

#Check for the name of the dataset and append it to a variable
filename <- list.files(directory, ".csv")

#Load data and append it to a variable
activity <- read.csv(file= filename, colClasses = "character", na.strings = "NA")

#Create directory for saving plots
figurespath <- paste0(directory, "/", "figures")
suppressWarnings(dir.create(path = figurespath))

#Define column classes
#Steps: Number of steps taking in a 5-minute interval
activity$steps <- as.numeric(activity$steps)

#Date : The date on which the measurement was taken in YYYY-MM-DD
# We will use *lubridate* package to transform date column from character to date

suppressMessages(library(lubridate))
activity$date <- ymd(activity$date)

#Interval : Identifier for the 5-minute interval in which measurement was taken
activity$interval <- as.numeric(activity$interval)

#Lets's check out data frame's information. 
str(activity)

```

## Phase II: Question #1- What is mean total number of steps taken per day?

In order to ask this question, we will create a subsequent table from the *activity*
dataset which contains the mean and the median for each day. *dplyr* package will be
used for summarization purposes. After the creation of the table, we will create a histogram
by using *ggplot2* package. At the end, we will report the daily max, mean and median of
steps taken. 

```{r}
library(ggplot2)
suppressMessages(library(dplyr)) 

spd <- suppressMessages(activity %>% group_by(date) %>% 
     summarize(sumSteps= sum(steps, na.rm = TRUE))) 

#As we can ignore the incomplete cases, let's remove them
spd <- spd[complete.cases(spd),]

#Let's check our table before creating a histogram
head(spd)

```

```{r}
#Create a histogram

ggplot(data=spd, aes(x=factor(date), y=sumSteps)) +
     geom_bar(stat="identity", fill="darkblue") +
     xlab("Dates")+
     ylab("Steps taken per day") +
     ggtitle("Steps Taken Per Day") +
     theme(axis.text.x = element_text(angle = 90, size = 5)) +
     theme(plot.title = element_text(hjust = 0.5))

#Save the plot
suppressMessages(ggsave("plot1.png", path = figurespath, device = "png")) 
```

Let's calculate the mean and median

```{r}
spd %>% summarize(meanSteps = mean(sumSteps, na.rm = TRUE), 
                  medianSteps = median(sumSteps, na.rm = TRUE))
```


## Phase III: Question #2: What is the average daily activity pattern?

First, group steps by interval and create the table using *ggplot2* package

```{r}
suppressMessages(stepsInterval <- suppressWarnings(activity %>% group_by(interval) %>% 
     summarize(average= mean(steps, na.rm = TRUE))))  

#Check the newly created dataset
head(stepsInterval)
```

```{r}
ggplot(data= stepsInterval, aes(x=interval, y= average)) +
     geom_line() + xlab("5-min Interval") + ylab("Steps Taken") +
     ggtitle("Average Steps Taken at Each 5-mins Interval") +
     theme(plot.title = element_text(hjust=0.5, face = "bold"))

#Save the plot
suppressMessages(ggsave("plot2.png", path = figurespath, device = "png"))
```

Let's calculate the "In which interval, the steps were at the maximum number on
average?"

```{r}
stepsInterval[which.max(stepsInterval$average),]

```

## Phase IV: Imputing Missing Values

1. Calculate the total number of rows with missing values

```{r}
incomplete_cases <- sum(!complete.cases(activity))
print( paste0("Number of rows with missing values are : ", as.character(incomplete_cases)))
```

2. Strategy for filling NAs

My strategy for filling the NAs in each column: Calculate the 5-mins interval for
each interval and append the calculated value to the NAs for the respective column.

```{r}
# First, check which columns has NA values

any(is.na(activity$steps))
any(is.na(activity$date))
any(is.na(activity$interval))
```

```{r}
# Calculte the average steps for each interval. We actually calculated this and
# stored in *stepsInterval*. Let's see it.

head(stepsInterval)
```
3. Creating the new dataset
As we know which values to append right now, let's create a new dataset with 
no missing values. We can use *ifelse()* function. We will test if it is NA, if yes,
append value from *stepsInterval* dataset if no, continue with existing.

```{r}
newactivity <- activity

newactivity$steps <- ifelse(is.na(newactivity$steps), stepsInterval$average,
                            newactivity$steps)

```

4. Number of steps taken each day

We will create an histogram for steps taken each day with this new dataset. We
will first create a table that contains the sum of steps for each day and create
the plot.

```{r}
suppressMessages(spd_new <- newactivity %>% group_by(date) %>% summarize(totalSteps = sum(steps)))

ggplot(data=spd_new, aes(x=date, y= totalSteps)) + 
     geom_bar(stat= "identity", fill= "darkblue")+
     xlab("Dates")+
     ylab("Number of Steps") +
     ggtitle("Steps Taken Per Day (NAs Imputed)") +
     theme(axis.text.x = element_text(angle = 90, size = 5)) +
     theme(plot.title = element_text(hjust = 0.5))

#Save the plot

suppressMessages(ggsave("plot3.png", path = figurespath, device = "png"))
```
5. Mean and median of steps

We have already created a dataset that contains steps taken in each day. We will
use that dataset to calculate the mean and median.

```{r}

spd_new %>% summarize(meanSteps = mean(totalSteps), medianSteps = median(totalSteps))

```

## Phase V: Question #3: Are there differences in activity patterns between
## weekdays and weekends?

First, we will create a new variable called "days". After that, we will create a
table that contains dates, days, steps per day variables. Lastly, we will create a
histogram in weekdays/weekends facet style. We will use *lubridate*, *dplyr* and
*ggplot2* packages. 

```{r}
#Create days variable

newactivity$days <- weekdays(newactivity$date, abbreviate = TRUE)

#Create a new variable for weekdays/weekends.

newactivity$part <- ifelse(newactivity$days %in% c("Cts", "Paz"), "Weekend", "Weekdays")

# Create a table that summarizes the results by grouping on dates and weekdays/weekends
suppressMessages(steps_part <- newactivity %>% 
     group_by(interval,part) %>% summarize(interval = interval,
                                            totalSteps = mean(steps),
                                            part = part)) 

#Draw plot
ggplot(data=steps_part, aes(x=interval, y=totalSteps)) +
     geom_line() +
     facet_grid(part ~ .) +
     xlab("5-min Interval") +
     ylab("Average Steps Taken") +
     ggtitle("Comparison of Average Steps Taken in 5- mins Interval \n between Weekdays and Weekands") +
     theme(plot.title = element_text(hjust = 0.5))

#Save Plot
suppressMessages(ggsave("plot4.png", path = figurespath, device = "png"))
```


## Phase VI: Knit the Document

```{r}
library(knitr)
library(rmarkdown)
```

After loading these packages, click the "Knit" option in the GUI system of RStudio.
It is in the toolbar of the left upper quadrant window.